%Format:
\documentclass[12pt,twoside,draft,en,bysec]{cmpart}
% Необов'язкові опції :
% draft - чернетка
% en, ua, ru - основна мова документа
% bysec - твердження, леми, означення і т.п. (але не теореми)
%         нумеруються в межах кожної \section окремо
% thmbysec - теореми теж нумеруються окремо для кожної \section
% win   - кодування кирилиці CP-1251 (Windows)
% koi   - кодування кирилиці KOI8_U  (Unix)
%
% Вже завантажено пакети :
% amsmath,amsfonts,amsthm,amssymb,latexsym
%
% Вже означено theorem-like оточення :
% {theorem}     - теорема
% {proposition} - твердження
% {lemma}       - лема
% {corollary}   - наслідок
% {remark}      - зауваження
% {definition}  - означення
% {example}     - приклад
% {examples}    - приклади
% а також їх аналоги з зірочкою без нумерації :
% {theorem*}, {proposition*} і т.п.


\author{Labachuk O.V., Zagorodnyuk A.V.}
\title{Multiplicative polynomial mappings on commutative Banach algebras}
\shorttitle{Multiplicative polynomials on commutative algebras}

\enabstract{Labachuk O.V., Zagorodnyuk A.V.}%
{Multiplicative polynomial mappings on commutative Banach algebras}%
{We consider the multiplicative polynomial mappings on commutative
algebras in this work. We call a multiplicative polynomial
trivial, if it can be represented as a product of characters. In
the paper we investigate the following question: does there exists
a nontrivial multiplicative polynomial functional on a commutative
algebra?}

\ruabstract{Лабачук О.В., Загороднюк А.В.}%
{Мультипликативные полиномиальные отображения на коммутативных банаховых алгебрах}%
{В этой работе мы рассматриваем мультипликативные полиномиальные
отображение на коммутативных алгебрах. Мультипликативный полином,
которой можна записать произведением характеров, назовем
тривиальным. Исследуем вопрос: существует ли не\-три\-виаль\-ный
мультипликативный полином на коммутативной алгебре?}

\uaabstract{Лабачук О.В., Загороднюк А.В.}%
{Мультиплікативні поліноміальні відображення на кому\-та\-тив\-них банахових алгебрах}%
{У цій роботі ми розглядаємо мультиплікативні поліноміальні
відображення на кому\-та\-тив\-них алгебрах. Мультиплікативний
поліном, що розкладається в добуток характерів,
на\-зи\-ва\-ти\-ме\-мо тривіальним. Досліджуємо питання: чи існує
нетривіальний
мультип\-лі\-ка\-тив\-ний поліном на комутативній алгебрі?}%


\subjclass{15A69, 46J20, 46G20} \UDC{517.98}
\keywords{multiplicative polynomial functional, characters,
commutative algebras}

\pyear{2012} % рік публікації
\volume{4}   % том
\issue{2}    % випуск
\pageno{284} % номер першої сторінки
\received{03.09.2012}
%\revised{?}
\contentslistfile{../total.lst}


\address{Vasyl Stefanyk Precarpathian National University,\\
Ivano-Frankivsk, Ukraine}
%\email{olabachuk@gmail.com}




\def\baselinestretch{1.1}

\begin{document}

\maketitle

%%% ----------------------------------------------------------------------

\section*{Introduction and definitions}

Let $X$ and $Y$ be linear spaces. A map $P_n: X\to Y$ is an \emph{
$n$ degree homogeneous polynomial} ($n$-homogeneous polynomial) if
there is an $n$-linear mapping $B_n: \underbrace{X\times
X\times\ldots \times X}_n \to Y$ such that
$P_n(x)=B_n(x,x,\ldots,x).$


Note that according to the polarization formula (see \cite{Mu}),
for every $n$-homogeneous po\-ly\-no\-mial $P_n$ there exists a
unique symmetric $n$-linear mapping $B_n,$ associated with $P_n.$

The polynomial mapping $P$ between algebras $X$ and $Y$ is called
\emph{multiplicative,} if $P(x_1\cdot x_2)=P(x_1)\cdot P(x_2)$ for
every $x_1,x_2 \in X.$ It is known that every multiplicative
polynomial functional is homogeneous (\cite{Z-p}).

Let us denote by $M(X)$ the set of characters (linear
multiplicative functionals) on the algebra $X.$ It is clear that
the product of characters
$\phi_1\cdot\phi_2\cdot\ldots\cdot\phi_n,$ $\phi_k\in M(X),$
$k=1,\ldots,n$ is a multiplicative polynomial functional.

We call a multiplicative polynomial \emph{trivial}, if it can be
represented as a product of characters.

There is an example of nontrivial multiplicative polynomial on
noncommutative algebra. Let us consider the algebra of square
$n\times n$ matrixes $M_n.$ The mapping $d(A)=\det(A)$ is a
multiplicative polynomial functional, but $d(A)$ is not a product
of characters because $M_n$ has no nonzero characters. So, it is
interesting to know: \emph{Does there exists a nontrivial
multiplicative polynomial functional on a commutative algebra?}





\section{Main results}

\begin{proposition}

If a commutative semi-simple Banach algebra $A$ without nilpotent
elements admits  a nontrivial multiplicative polynomial, then the
algebra $P(\mathbb{C}^m)$ of polynomials of $m$ variables admits a
nontrivial multiplicative polynomial for some $m.$

\end{proposition}

\proof Let $P$ be a nontrivial multiplicative polynomial map.
Without loss of the ge\-ne\-ra\-li\-ty we can assume that $P$ is
irreducible. Then there exists a finite dimensional subspace $V$
in $ A$ such that the restriction of $P$ onto $V$ is irreducible
\cite{Zag}.

Let us consider a subalgebra $A_0$  generated by elements in $V$
and the unity of $A.$ This subalgebra is finitely generated and
has no nilpotent elements. It is well known in Algebraic Geometry
that such algebra is isomorphic to a ring of all polynomials on an
algebraic variety. That is, $A_0$ is isomorphic to
$P(\mathbb{C}^m)/I$ for some $m$ where $I$ is an ideal in
$P(\mathbb{C}^m).$ Let $T: P(\mathbb{C}^m)\rightarrow
P(\mathbb{C}^m)/I$ be the factor map and $P_0$ be the restriction
of $P$ onto $A_0.$ Then $P_0\circ T$ is a required irreducible
nontrivial multiplicative polynomial on $P(\mathbb{C}^m).$
\hfill$\square$

So, it can be useful to investigate the algebra $P(\mathbb{C}^m)$
of polynomials of $m$ variables.


\begin{theorem}

Every multiplicative  $n$ degree polynomial functional  can be
represented as a product of characters on the algebra
$P(\mathbb{C})$ of polynomials of one complex variable.

\end{theorem}

\begin{proof}
Let $D_n: P(\mathbb{C}) \rightarrow \mathbb{C}$ be a
multiplicative $n$ degree polynomial functional. $D_n$ is a
homogeneous polynomial map. According to \cite{Z-p} there is a
character $d_n$ on the symmetric tensor product
$\underbrace{P(\mathbb{C})\otimes_s\ldots\otimes_sP(\mathbb{C})}_n=\otimes^n_sP(\mathbb{C})$
such that $D_n(p)=d_n(\otimes^np)$ for every $p\in P(\mathbb{C}).$

Let us describe the algebra $\otimes^n_sP(\mathbb{C}).$ Every
element of this algebra can be presented by
$$\sum_i
\underbrace{p_i\otimes\ldots\otimes p_i}_n,$$ where $p_i\in
P(\mathbb{C}).$ Every element like $\otimes^np_i$ is a polynomial
of $n$ variables $p_i(x_1)\cdot\ldots\cdot p_i(x_n).$ So, algebra
$\otimes^n_sP(\mathbb{C})$ is isomorphic to the algebra of
 polynomials of $n$ variables, that are symmetric about the permutation of this variables.
 Let us denote it by
$P_s(\mathbb{C}^n).$

It is well known that for every symmetric polynomial
$p(x_1,\ldots,x_n)$ there exists a polynomial $q,$ such that
$$p(x_1,\ldots,x_n)=q(G_1(x_1,\ldots,x_n),\ldots,G_n(x_1,\ldots,x_n)),$$
 where $$G_1(x_1,\ldots,x_n)=x_1+\ldots+x_n,$$
$$G_2(x_1,\ldots,x_n)=x_1x_2+x_1x_3+\ldots+x_{n-1}x_n,$$
$$\ldots    \ldots     \ldots$$
$$G_n=x_1x_2\cdot\ldots\cdot x_n,$$
that is $G_i,$ $i=1,\ldots,n$ are the elementary symmetric
polynomials. The mapping $p \mapsto q$ is an isomorphism from the
algebra $P_s(\mathbb{C}^n)$ of symmetric polynomials onto the
algebra of all polynomials $P(\mathbb{C}^n).$ Every character of
algebra $P(\mathbb{C}^n)$ is an evaluation of some point
$\mathbb{C}^n.$ So, there is a correspondence between characters
$d_n$ on $\otimes^nP(\mathbb{C})$ and some characters $\varphi$ on
$P(\mathbb{C}^n),$ which is an evaluation of polynomial $q$ at
some point $(\alpha_1, \ldots, \alpha_n)\in \mathbb{C}^n.$ That
is,
$$d_n(\otimes^n p)=d_n(p(x_1)\cdot\ldots \cdot p(x_n))=\varphi(q(x_1,\ldots,x_n))=q(\alpha_1, \ldots, \alpha_n),$$ where
$p(x_1)\cdot\ldots \cdot
p(x_n)=q(G_1(x_1,\ldots,x_n),\ldots,G_n(x_1,\ldots,x_n)).$ Let
$x_1^0,\ldots,x_n^0$ be the solutions of system  $$ \left\{
\begin{array}{ll} x_1+x_2+\ldots+x_n=\alpha_1; &  \\ x_1x_2+x_1x_3+\ldots+x_{n-1}x_n=\alpha_2; &
\\ \ldots \ldots \ldots \ldots &
\\ x_1x_2\cdot\ldots\cdot x_n=\alpha_n
\end{array} \right..$$ According to the Viet theorem,
$x_1^0,\ldots,x_n^0$ are the solutions of the equation
$$x^n-\alpha_1x^{n-1}+\alpha_2x^{n-2}+\ldots+(-1)^n\alpha_n=0.$$

Then $D_n(p)=q(\alpha_1, \ldots, \alpha_n)=p(x_1^0)\cdot\ldots
\cdot p(x_n^0),$ that was needed to show.
\end{proof}

To prove of this theorem we  use the existence of a homomorphism
from $\otimes^nP(\mathbb{C})=P(^n\mathbb{C})$ onto
$\otimes^n_sP(\mathbb{C}).$ It is easy to show that this condition
is sufficient in the general case.

\begin{theorem}
If there exists a surjective homomorphism  $\phi: \otimes^n A
\rightarrow \otimes^n_{s}A,$  then every $n$-homogeneous
multiplicative polynomial on algebra $A$ can be represented as a
product of characters.
\end{theorem}

\begin{proof} Suppose that there exists the surjective homomorphism $\phi$
from $\otimes^n A$ onto $\otimes^n_{s}A.$ For a given
$n$-homogeneous multiplicative polynomial $D(x), x\in A,$ it is
defined a character $d$ on $\otimes^n_{s}A$ such that
$$d(\otimes^np)=D(x).$$
Thus
$$\otimes^n A
\rightarrow \otimes^n_{s}A \rightarrow \mathbb{C}$$

Let $p\in \otimes^n_{s}A$ and $u\in \varphi^{-1}(p),$ $u\in
\otimes^n A.$ Then $d\circ \varphi$ is a character on $\otimes^n
A$ and $d\circ \varphi(u)=d(p).$ According to \cite{je}, every
character of $\otimes^n A$ is of the form
$$d\circ \varphi(a\otimes\ldots\otimes a)=\psi_1(a)\ldots\psi_n(a),$$
$a\in A$ for some characters $\psi_1,\ldots,\psi_n$ on $A.$ So if
$p=a\otimes\ldots\otimes a,$ then $$D(a)=d(p)=d\circ
\varphi(u)=\psi_1(a)\ldots\psi_n(a),$$ that is, $D$ is trivial.
\end{proof}


\begin{corollary}

Let $A$ be the completion of algebra $P(\mathbb{C})$ in some
locally convex metrizabled topology $\tau,$ such that $(A, \tau)$
is a topology algebra. Then every multiplicative polynomial by $n$
degree on $A$ can be presented as a product of characters.

\end{corollary}

\begin{theorem}
Every multiplicative polynomial of second degree is trivial on the
algebra $P(\mathbb{C}^2)$ of polynomials of two variables.

\end{theorem}

\begin{proof}
We use a similar idea that in proof of  Theorem 1. Let $D_2:
P(\mathbb{C}^2) \rightarrow \mathbb{C}$ be a homogeneous
 multiplicative polynomial by second degree. Due to
\cite{Z-p}, there exists the character $d$ on the symmetric tensor
product $P(\mathbb{C}^2)\otimes_sP(\mathbb{C}^2)$ such that
$D_2(p)=d_2(p\otimes p)$ for any $p\in P(\mathbb{C}^2).$

The algebra $P(\mathbb{C}^2)\otimes_sP(\mathbb{C}^2)$ is
isomorphic to an algebra of four variables polynomials generated
by $p(x_1,y_1)p(x_2,y_2)$, which are symmetric about the
permutations of pairs  $x_1,y_1$ and $x_2,y_2$ simultaneously.
These polynomials are called block-symmetric polynomials.

A polynomial wich is symmetric with respect to the permutation of
pairs
 $x_1,y_1$ and
$x_2,y_2,$ can be presented by polynomials $R_1=x_1+x_2,$
$R_2=x_1x_2,$ $R_3=y_1+y_2,$ $R_4=y_1y_2,$ $R_5=x_1y_2+x_2y_1,$
that is $p(x_1,y_1)p(x_2,y_2)=q(R_1,R_2,R_3,R_4,R_5).$ But the
polynomials $R_1,R_2,R_3,R_4,R_5$ are algebraically depending and
this dependence can be wrote by the following formula:
$$R_5^2-R_1R_3R_5+R_1^2R_4-4R_2R_4+R_3^2R_2=0.\eqno(1)$$

Thus the polynomial $q$ is an element of factor-algebra which is
generated by polynomial $(1)$ and it is determined by this
polynomial zeros. The character $d$ on
$P(\mathbb{C}^2)\otimes_sP(\mathbb{C}^2)$ is a value of polynomial
$q$ at the point $(\alpha_1, \alpha_2,\alpha_3,\alpha_4,
\alpha_5),$ which is the solution of the equation $(1),$ that is
$$d(p\otimes p)=d(p(x_1,y_1)p(x_2,y_2))=q(\alpha_1, \alpha_2,\alpha_3,\alpha_4, \alpha_5),$$
where $p(x_1,y_1)p(x_2,y_2)=q(R_1,R_2,R_3,R_4,R_5).$ Let
$x_1^0,x_2^0,y_1^0,y_2^0$ be the solution of the equations system
$$ \left\{
\begin{array}{ll} x_1+x_2=\alpha_1; &  \\ x_1x_2=\alpha_2; &
\\ y_1+y_2=\alpha_3;  &
\\ y_1y_2=\alpha_4; &
\\ x_1y_2+x_2y_1=\alpha_5
\end{array} \right..$$ According to the Viet theorem
the pairs  $x_1^0,x_2^0$  and $y_1^0,y_2^0$ are the solution of
the equations
$$x^2-\alpha_1x+\alpha_2=0,$$
$$y^2-\alpha_3x+\alpha_4=0$$
respectively. Now we need to put $\alpha_5=x_1^0y_2^0+x_2^0y_1^0.$
Then $D_2(p)=q(\alpha_1, \alpha_2,\alpha_3,\alpha_4,
\alpha_5)=p(x_1^0,y_1^0)p(x_2^0,y_2^0),$ that was needed to show.
\end{proof}

Using similar methods and recent results on block-symmetric
polynomials \cite{bl-sym} it is possible to prove an analogue of
Theorem 3 for $n$ degree multiplicative polynomials on
$P(\mathbb{C}^2).$




%% Оформлення літератури
\begin{thebibliography}{99}


\bibitem{bl-sym}
Загороднюк А.В., Кравцiв В.В. \emph{Симетричнi полiноми на
добутках банахових просторiв}
// Карпат. мат. публ., --- 2010. --- Т.2, №1. --- C. 59--71.

\bibitem{Kol}  Колмогоров А.,
 Фомин С. Елементы теории функций  и функционального анализа.
--- М.: Наука, 1976. --- 543 с.

\bibitem{Ga}
 Gamelin T.W. Analytic functions on Banach spaces in Complex
Function
  Theory, Ed. Gauthier and Sabidussi,  Kluwer Academic
  Publishers, Amsterdam, 1994.

\bibitem{je}
 Lamadrid J.~G. \textit{Uniform cross norms and tensor products of
Banach algebras,}  Bull. Amer. Math. Soc., \textbf{69}, 6 (1963),
797--803.

\bibitem{Mu}
Mujica J. Complex Analysis in Banach Spaces, North-Holland,
  Amsterdam, New York, Oxford, 1986.

\bibitem{Ru}  Rudin W. Functional Analysis, McGraw-Hill, New
York, 1973.

\bibitem{Z-p}  Zagorodnyuk A.V. {\it Multiplicative polynomial operators on
topological
algebras,} Contemporary Mathematics, \textbf{232} (1999),
357--361.


\bibitem{Zag}  Zagorodnyuk A.V. {\it On polynomial orthogonality on
Banach spaces,} Math. Studii, \textbf{14}, 2 (2000),
189--192.






\end{thebibliography}

\end{document}

\begin{thebibliography}{9}

  \bibitem{k1} Иосида К. Функциональный анализ. --- М.: Мир, 1967. --- 624 с.

  \bibitem{k2} Малицька Г.П. \emph{Системи рiвнянь типу Колмогорова} // Укр.
мат. журн. --- 2009. --- Т.12, №3. --- С. 1650--1563.

  \bibitem{k3} Dineen S. Complex analysis on infinite-dimensional spaces,
Springer Monographs in Mathematics,
Springer-Verlag London, Ltd., London, 1999.


\end{thebibliography}

%\end{document}
